\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-09-15 submission/manuscript.Rmd
Update manuscript.Rmd

<!-- Goals and hypotheses -->
This study aims primarily at exploring the effect of face attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between the perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely believed to be real. Moreover, we expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the role of dispositional traits, such as personality and attitude towards AI, characteristics on inter-individual simulation monitoring tendencies.

@@

109 photos of neutral-expression faces of real individuals were retrieved from the American Multiracial Face Database (AMFD, [@chen2021broadening]). In the first part of the study, participants answered a series of personality questionnaires. This includes the *Mini-IPIP6* [24 items, @sibley2011], to measure 6 "normal" personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility); the *FFNI-BF* [30 items, @jauk2022validation] to measure 9 facets of narcissism (acclaim-seeking, distrust, entitlement, exploitativeness, indifference, lack of empathy, manipulativeness, need for admiration and thrill-seeking); the *R-GPTS* [18 items, @freeman2021revised], to measure 2 sub-scales of paranoid thinking (persecution and reference ideation); the *IUS-12* [12 items, @carleton2007fearing], to assess beliefs about uncertainty and the future; the *SIAS-6* and the *SPS-6* [6 items each, @peters2012development], to assess social anxiety levels as well as a 10-item scale that measures participants general attitudes towards artificial intelligence (of which 5 items were adapted from the *GAAIS* [@schepman2020initial]).

In the second part of this study, the set of facial images retrieved from the AMFD were presented to the participants for 500ms each, in a randomized order. Following each facial display, participants were asked to rate each face on its **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") using a visual analog scale.

In the last part of the study, participants were informed that half of the facial images previously seen were AI-generated. The same set of facial images displayed before were then presented for a second set of trials in a new randomized order. As per the first set, each facial stimuli was presented for 500ms. After each display, participants were asked to judge whether they thought the face they saw was real (or fake) using a visual analog scale.

This experiment was implemented using *jsPsych* [@de2015jspsych], and the full set of instructions is available in the experiment code.

@@

103 participants were recruited via *Prolific* ([www.prolificacademic.co.uk](www.prolificacademic.co.uk)), a crowd-sourcing platform providing high data quality [@peer2022]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 50 minutes to finish. Demographic variables (age, gender, and ethnicity) were self-reported on a voluntary basis.

Upon inspection of stimuli ratings, attention check responses and total duration spent to complete the personality questionnaires, 3 participants were removed for spending an implausibly short time to finish the questionnaires and having very low correlations between their stimuli ratings and that of the normative distribution.

The final sample included 100 participants (Mean age = 27.9, SD = 8.5, range: [19, 66]; Sex: 48% females, 52% males).

>>>>

<!-- Goals and hypotheses -->
This study aims primarily at exploring the effect of face attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between the perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely believed to be real. Moreover, we expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the role of dispositional traits, such as personality and attitude towards AI, characteristics on inter-individual simulation monitoring tendencies. Note that the discriminative accuracy between "true" photos and "true" artificially-generated images is not relevant for this study, which focuses on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

In the first part of the study, participants answered a series of personality questionnaires, including the *Mini-IPIP6* [24 items, @sibley2011] measuring 6 personality traits, the *FFNI-BF* [30 items, @jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS* [18 items, @freeman2021revised] measuring 2 dimensions related to paranoid thinking; the *IUS-12* [12 items, @carleton2007fearing] measuring intolerance to uncertainty, the *SIAS-6* and the *SPS-6* [6 items each, @peters2012development], assessing social anxiety levels. Finally, we created 5 items pertaining to expectations about AI-generated images technology (**TODO: write here some of the questions**). To lower their saliency and possibly prime the subjects about the task, we mixed these items with 5 items of the attitudes towards AI scale [*GAAIS*, @schepman2020initial]. 3 attention check questions were embedded in the questionnaires.

In the second part of this study, 109 photos of neutral-expression faces of real individuals from the American Multiracial Face Database (AMFD, [@chen2021broadening]) were presented to the participants for 500ms each, in a randomized order. Following each stimulus, ratings of **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") were collected using visual analog scales.

In the last part of the study, participants were informed that about half of the face images previously seen were AI-generated (the instructions used a cover story mentioning that the research was aimed at validating a new face generation algorithm). The same set of stimuli was displayed again for 500 ms in a new randomized order. This time, after each display, participants were asked to express their belief regarding the nature of the stimulus using a visual analog scales (with *Fake* and *Real* as the two extremes). The study was implemented using *jsPsych* [@de2015jspsych], and the full set of instructions is available in the experiment code.

@@

One hundred and three participants were recruited via *Prolific*, the the crowd-sourcing platform providing the best data quality [@peer2022]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 45 minutes to finish. Demographic variables (age, gender, sexual orientation, education and ethnicity) were self-reported on a voluntary basis.

We excluded 3 participants that failed 2 (out of 3) or more attention check questions. The final sample included 100 participants (Mean age = 27.9, SD = 8.5, range: [19, 66]; Sex: 48% females, 52% males).
